\documentclass[11pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{array}
\usepackage{booktabs}
\usepackage{float}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}

\numberwithin{equation}{section}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\TT}{\mathbb{T}}
\newcommand{\Fold}{\mathrm{Fold}}
\newcommand{\leb}{\mathrm{Leb}}
\newcommand{\pref}{\mathrm{pref}}
\DeclareMathOperator{\TV}{TV}
\DeclareMathOperator{\KL}{KL}
\DeclareMathOperator{\supp}{supp}

\title{Zeckendorf folds, Sturmian rigidity, and Parry divergence on the golden-mean shift}
\author{Haobo Ma\thanks{AELF PTE LTD., \#14-02, Marina Bay Financial Centre Tower 1, 8 Marina Blvd, Singapore 018981, Singapore. \texttt{auric@aelf.io}.}
\and
Wenlin Zhang\thanks{National University of Singapore, Singapore. Corresponding author. \texttt{e1327962@u.nus.edu}.}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We study the canonical Zeckendorf fold $\Fold_m:\{0,1\}^m\to X_m$, where
$X_m$ is the length-$m$ language of the golden-mean shift. On the canonical
Sturmian slice $\beta=\alpha$, the restriction of $\Fold_m$ to realized
rotation factors is bijective and induces the classical higher-block
presentation of the Sturmian subshift. Using the endpoint collapse of
golden-mean Parry cylinders, we derive an exact formula for
$\KL(\nu\|\pi_m)$, sharp support-constrained minima for all R\'enyi orders,
and, for $m\ge 6$, the exact optimal injective placement of the realized
Sturmian law into $X_m$. For bounded-type slopes this yields
$$
D_q(\pi_m^{\mathrm{fold}}\|\pi_m)
=m\log\varphi-\log m+O_\alpha(1),
$$
uniformly in $q$. We further show that this finite-$m$ formula calibrates
the universal injective-placement barrier for lossless coarse-grainings into
primitive shifts of finite type, and that the Fibonacci weighting is uniquely
characterized by a natural contiguous numeration axiom. A short final section
records the conjugacy-invariant transfer of the construction to irrational
circle diffeomorphisms and includes one numerical illustration.
\end{abstract}

\medskip
\noindent\textbf{Keywords.} Sturmian subshift, Zeckendorf numeration,
golden-mean shift, Parry measure, R\'enyi divergence, shifts of finite type

\medskip
\noindent\textbf{2020 Mathematics Subject Classification.}
37B10 (primary), 37A55, 11B39, 94A17

\input{sec1_introduction}
\input{sec2_fold_language}
\input{sec3_sturmian_rigidity}
\input{sec4_parry_divergence}
\input{sec5_injective_placement}
\input{sec5b_primitive_sft}
\input{sec6_conjugacy}
\input{sec7_concluding}

\appendix
\input{appA_discrepancy}
\input{appB_fiber}
\input{appC_normalization}

\bibliographystyle{unsrt}
\bibliography{../../2026_golden_ratio_driven_scan_projection_generation_recursive_emergence/references,../../2026_golden_ratio_driven_scan_projection_generation_recursive_emergence/references_fibonacci,../../2026_golden_ratio_driven_scan_projection_generation_recursive_emergence/references_internal_2025}

\end{document}
