\section{Introduction}\label{sec:introduction}

This paper studies the canonical Zeckendorf fold, a deterministic map
$\Fold_m:\{0,1\}^m\to X_m$ that sends binary words to the length-$m$
language $X_m$ of the golden-mean shift by Fibonacci-weighted evaluation and
Zeckendorf normalization. Three structural phenomena are proved.

First, on the Sturmian slice $\beta=\alpha$, the fold is bijective on
realized rotation factors and the induced sequence map is topologically
conjugate to the classical higher-block presentation of the Sturmian
subshift. Minimality, unique ergodicity, zero entropy, and linear word
complexity are therefore preserved exactly.

Second, the endpoint collapse of golden-mean Parry cylinders yields a closed
formula for $\KL(\nu\|\pi_m)$ and, more generally, sharp
support-constrained Parry divergence minima for all R\'enyi orders. The
formula depends only on the endpoints of the admissible words, and its
universality over $q\in[0,\infty]$ is a consequence of the scalar
endpoint statistic peculiar to the golden-mean transition matrix. For
$m\ge 6$, the exact optimal injective placement of the realized Sturmian
law into $X_m$ is determined: the infimum over all injections
$\iota:S_m(\alpha)\hookrightarrow X_m$ equals
\[
L_q(m)=(m-1)\log\varphi-\log\pi(0)-H_q(\mu_m),
\]
and every injection whose image lies in the heaviest endpoint class
$X_m^{00}$ attains it. Moreover, the full $O(1)$ defect between the
canonical folded law and the injective optimum is shown to be an exact
escort-pressure of a boundary cocycle
(Theorem~\ref{thm:full-renyi-defect-formula}), not an entropy residual.

Third, for bounded-type slopes the folded law satisfies
\[
D_q(\pi_m^{\mathrm{fold}}\|\pi_m)
=m\log\varphi-\log m+O_\alpha(1),
\qquad q\in[0,\infty],
\]
and the canonical fold remains within a bounded additive distance of the
injective optimum. This finite-$m$ formula calibrates the universal
injective-placement barrier for primitive shifts of finite type
(Theorem~\ref{thm:primitive-sft-injective-barrier}): any deterministic map
that is injective on the realized source retains the order-uniform
divergence rate $D_q^{\mathrm{opt}}=mh-H_q(\mu_m)+O(1)$. The golden-mean
benchmark is exceptional because this universal envelope becomes finite-$m$
exact, the optimizer is explicit, and the actual canonical fold remains
within a bounded additive distance of that optimizer. Moreover, the
Fibonacci weighting is proved to be the unique choice satisfying a natural
contiguous numeration axiom (Theorem~\ref{thm:fibonacci-weight-uniqueness}).

The ingredients are individually classical. Sturmian codings, factor
complexity, and factor frequencies go back to
\cite{MorseHedlund1940,Slater1950DistributionIntegers,Slater1967GapsSteps}
and fit naturally into the substitution and symbolic-dynamical framework
\cite{LindMarcus1995,QueffelecSubstitution,Fogg2002}. The Parry measure and
the thermodynamic formalism of shifts of finite type are standard
\cite{LindMarcus1995,WaltersErgodicTheory,ParryPollicott1990Zeta}. Fibonacci
numeration and finite-state normalization are likewise classical
\cite{Zeckendorf1972,Frougny1991FibonacciFiniteAutomata,Frougny1992NumbersAutomata},
with related Sturmian and Ostrowski numeration viewpoints in
\cite{Berthe2001Ostrowski,BaranwalaSchaefferShallit2021OstrowskiAutomatic}.
What is specific to the present construction is that these ingredients
combine cleanly: the fold is dynamically lossless on the realized Sturmian
source, while the comparison with the Parry measure is explicit enough to
yield exact finite-$m$ optimization statements. On the numeration side,
Ostrowski and related continued-fraction systems
\cite{Berthe2001Ostrowski,BaranwalaSchaefferShallit2021OstrowskiAutomatic}
provide arithmetic codings with rich dynamical content; the present paper
uses the golden branch of that numeration-theoretic viewpoint as an exactly
analyzable coarse-graining whose information geometry relative to the Parry
law can be solved sharply.

For primitive one-step shifts of finite type, support-constrained R\'enyi
optimization remains finite-$m$ exact at the endpoint-class level
(Theorem~\ref{thm:primitive-sft-endpoint-envelope}). The golden-mean case
is exceptional only because its endpoint spectrum collapses to three
Fibonacci classes, so the general endpoint envelope closes in elementary
form. This upgrades the role of the golden-mean setting from a single
benchmark to the most explicit instance of a general endpoint-class theory
for primitive shifts. By topological conjugacy, the entire construction
transfers verbatim to nonlinear circle diffeomorphisms with irrational
rotation number.

The paper is organized as follows. Section~\ref{sec:fold-language} introduces
the canonical Zeckendorf fold, the golden-mean language, and the Parry
reference measure. Section~\ref{sec:sturmian-rigidity} proves the
realized-support bijection and the higher-block conjugacy on the Sturmian
source. Section~\ref{sec:parry-divergence} derives the exact Parry divergence
identity, the sharp support barriers, and the endpoint-class optimization.
Section~\ref{sec:injective-placement} establishes the optimal injective
relabeling, the bounded-type asymptotics, and the primitive-SFT extension.
Section~\ref{sec:conjugacy} records the conjugacy-invariant transfer to circle
diffeomorphisms and one numerical illustration. Section~\ref{sec:concluding}
collects concluding remarks. Three appendices supply auxiliary discrepancy and
empirical entropy bounds (Appendix~\ref{app:discrepancy}), fiber-uniform lifts
and I-projections (Appendix~\ref{app:fiber-lift}), and the local Zeckendorf
normalization together with the uniqueness of the Fibonacci weighting
(Appendix~\ref{app:canonical-fold}).
