\section{Sturmian Rigidity and Higher-Block Conjugacy}
\label{sec:sturmian-rigidity}

In the canonical Sturmian slice, the fold does not merge the words that are
actually realized by the rotation. The reason is arithmetic: along the
Sturmian cylinder partition, the Fibonacci values of the factors remain
distinct modulo \(F_{m+2}\).

\begin{theorem}[Sturmian no-collision rigidity]\label{thm:sturmian-no-collision-rigidity}
Assume \(\beta=\alpha\), and define
\[
S_m(\alpha):=\supp(\mu_m)\subset\Omega_m,
\qquad
Y_m:=\Fold_m(S_m(\alpha))\subset X_m.
\]
Then the restriction
\[
\Fold_m\!\upharpoonright_{S_m(\alpha)}:S_m(\alpha)\to Y_m
\]
is a bijection.

More precisely, let
\[
\Xi_m:=\{(1-j)\alpha \bmod 1:\ 0\le j\le m\}\subset\TT,
\]
let \(I_0,\dots,I_m\) be the open atoms of \(\TT\setminus \Xi_m\) in
counterclockwise order, and choose \(x_\ell\in I_\ell\). Writing
\[
u^{(\ell)}:=\omega_m(x_\ell;0)\in S_m(\alpha),
\qquad
V_\ell:=N(u^{(\ell)}),
\]
the cyclic jump multiset
\[
\Delta_\ell:=V_{\ell+1}-V_\ell
\qquad (\ell\!\!\!\!\pmod{m+1})
\]
consists of exactly one positive jump \(+F_{m+1}\) and the \(m\) negative
jumps
\[
-1,\ -F_1,\ -F_2,\ \dots,\ -F_{m-1},
\]
counted with multiplicity. After a cyclic relabeling that starts immediately to
the right of the unique positive jump, the values \(V_\ell\) form a strictly
decreasing integer chain. In particular, the residues \(V_\ell\bmod F_{m+2}\)
are pairwise distinct.
\end{theorem}

\begin{proof}
The proof has two steps: compute the Fibonacci jump produced by crossing each
Sturmian boundary atom, and then observe that after the unique positive jump
the remaining cycle is a strictly decreasing chain whose total drop is smaller
than \(F_{m+2}\), so residue collisions cannot occur. The boundary bookkeeping
is the same one that underlies the standard factor partition of Sturmian words;
see \cite{LindMarcus1995,Fogg2002,Berthe2001Ostrowski}.

For \(0\le k\le m-1\), the \(k\)-th symbol of \(\omega_m(x;0)\) is
\[
\mathbf{1}_{[0,\alpha)}(x+k\alpha),
\]
so its discontinuities occur at the two points \(-k\alpha\) and
\(\alpha-k\alpha=(1-k)\alpha\) on \(\TT\). The union of all such discontinuity
points is exactly \(\Xi_m\). Hence \(\omega_m(x;0)\) is constant on each atom
\(I_\ell\), and \(S_m(\alpha)=\{u^{(\ell)}:0\le \ell\le m\}\).

Fix a boundary point \(\xi_j:=(1-j)\alpha\in\Xi_m\).
If \(1\le j\le m-1\), then crossing \(\xi_j\) from left to right changes the
symbols in positions \(j-1\) and \(j\) from \(01\) to \(10\), while all other
positions stay fixed: among the translated points \(x+k\alpha\), only the two
with \(k=j-1\) and \(k=j\) meet the coding boundary at the instant \(x=\xi_j\).
Therefore the Fibonacci value jump is
\[
F_{j+1}-F_{j+2}=-F_j.
\]
At \(\xi_0=\alpha\), only the first symbol changes, from \(1\) to \(0\), so
the jump is \(-F_1=-1\). At \(\xi_m=(1-m)\alpha\), only the last symbol
changes, from \(0\) to \(1\), so the jump is \(+F_{m+1}\). This proves the
stated jump multiset.

Now start with the atom immediately to the right of \(\xi_m\). Along the
remaining cycle, every subsequent jump is negative, so the corresponding values
form a strictly decreasing integer chain. Since
\[
1+\sum_{j=1}^{m-1}F_j=F_{m+1},
\]
the total drop before the cycle closes is exactly \(F_{m+1}<F_{m+2}\). Hence
no two values in this chain can be congruent modulo \(F_{m+2}\).
Proposition~\ref{prop:fold-residue-characterization} shows that
\(\Fold_m\) is constant on residue classes modulo \(F_{m+2}\), so its
restriction to \(S_m(\alpha)\) is injective. Bijectivity onto \(Y_m\) is then
tautological.
\end{proof}

\begin{remark}[What the rigidity proof uses]
\label{rem:sturmian-rigidity-special}
The proof uses three ingredients that are all special to the canonical slice
\(\beta=\alpha\): the Sturmian partition of the torus into exactly \(m+1\)
atoms, the explicit Fibonacci weighting in \(N(\cdot)\), and the residue
characterization of the canonical fold modulo \(F_{m+2}\). The argument is
therefore more rigid than a generic coding proof, which is why an analogous
jumps-and-residues theorem is not claimed for non-canonical folds.
\end{remark}

\begin{corollary}[Exact relabeling on the realized Sturmian language]\label{cor:sturmian-exact-relabeling}
Assume \(\beta=\alpha\), and let
\[
B_m:=\Fold_m\!\upharpoonright_{S_m(\alpha)}:S_m(\alpha)\to Y_m.
\]
Then \(B_m\) is a bijection, and for every \(N\ge 1\),
\[
\pi_m^{\mathrm{fold}}=(B_m)_*\mu_m,
\qquad
\hat\pi_{m,N}=(B_m)_*\hat\mu_{m,N}.
\]
Consequently,
\[
\TV(\hat\pi_{m,N},\pi_m^{\mathrm{fold}})
=
\TV(\hat\mu_{m,N},\mu_m),
\]
\[
\KL(\hat\pi_{m,N}\|\pi_m^{\mathrm{fold}})
=
\KL(\hat\mu_{m,N}\|\mu_m),
\]
and, writing \(H_q\) for the R\'enyi entropy of order \(q\), for every
\(q\in[0,\infty]\),
\[
H_q(\hat\pi_{m,N})=H_q(\hat\mu_{m,N}),
\qquad
H_q(\pi_m^{\mathrm{fold}})=H_q(\mu_m).
\]
\end{corollary}

\begin{proof}
The pushforward identities follow from Theorem~\ref{thm:sturmian-no-collision-rigidity}.
Since \(B_m\) is a bijection between the realized supports, total variation,
relative entropy, and all R\'enyi entropies are invariant under the relabeling
\(B_m\).
\end{proof}

\subsection{Higher-Block Conjugacy on the Sturmian Source}

\begin{theorem}[Higher-block conjugacy on the Sturmian source]
\label{thm:sequence-level-transparency}
Assume \(\beta=\alpha\), let \(X_\alpha\subset\{0,1\}^{\mathbb{Z}}\) be the
Sturmian subshift generated by the irrational rotation coding, and fix
\(m\ge 1\). Define
\[
(\Phi_m(s))_t:=\Fold_m(s_t s_{t+1}\cdots s_{t+m-1}),
\qquad t\in\mathbb{Z}.
\]
Then \(\Phi_m:X_\alpha\to \mathcal{Y}_m:=\Phi_m(X_\alpha)\) is a topological
conjugacy. Equivalently, \(\mathcal{Y}_m\) is a letter relabeling of the
classical \(m\)-block presentation of \(X_\alpha\).

In particular, \(\mathcal{Y}_m\) is minimal and uniquely ergodic,
\[
h_{\mathrm{top}}(\mathcal{Y}_m)=0,
\]
and, writing \(p_Z(n)\) for the word complexity of a subshift \(Z\), one has
\[
p_{\mathcal{Y}_m}(n)=p_{X_\alpha}(n+m-1)=n+m
\qquad (n\ge 1).
\]
\end{theorem}

\begin{proof}
Let \(\Psi_m:X_\alpha\to S_m(\alpha)^{\mathbb{Z}}\) be the classical higher-block
code
\[
(\Psi_m(s))_t:=s_t s_{t+1}\cdots s_{t+m-1}.
\]
Standard higher-block theory gives that \(\Psi_m\) is a topological conjugacy
from \(X_\alpha\) onto its \(m\)-block presentation
\cite{LindMarcus1995}. By Corollary~\ref{cor:sturmian-exact-relabeling},
\[
B_m:=\Fold_m\!\upharpoonright_{S_m(\alpha)}:S_m(\alpha)\to Y_m
\]
is a bijection. Hence the coordinatewise relabeling
\[
(\widetilde B_m(z))_t:=B_m(z_t),
\qquad z\in \Psi_m(X_\alpha),
\]
is a topological conjugacy from the \(m\)-block presentation onto
\(\mathcal{Y}_m\).
Because
\[
\Phi_m=\widetilde B_m\circ \Psi_m,
\]
\(\Phi_m\) is itself a topological conjugacy.

Minimality and unique ergodicity are preserved under conjugacy, and the
irrational rotation coding is classically minimal and uniquely ergodic. The
same is therefore true of \(\mathcal{Y}_m\). Likewise, topological entropy is
preserved, so \(h_{\mathrm{top}}(\mathcal{Y}_m)=h_{\mathrm{top}}(X_\alpha)=0\).

A word of length \(n\) in the \(m\)-block presentation corresponds to a word of
length \(n+m-1\) in \(X_\alpha\), and the relabeling \(\widetilde B_m\) does
not change word counts. Therefore
\[
p_{\mathcal{Y}_m}(n)=p_{X_\alpha}(n+m-1).
\]
Since \(X_\alpha\) is Sturmian, \(p_{X_\alpha}(k)=k+1\) for every \(k\ge 1\),
which gives \(p_{\mathcal{Y}_m}(n)=n+m\).
\end{proof}

\begin{corollary}[Pure degeneracy ledger on the Sturmian slice]\label{cor:sturmian-pure-degeneracy-ledger}
Assume \(\beta=\alpha\), and let \(\mu_m^e:=Q_m\pi_m^{\mathrm{fold}}\). Then
\[
\KL(\mu_m\|\mu_m^e)
=
\mathbb{E}_{\pi_m^{\mathrm{fold}}}\bigl[\log d_m(Y)\bigr].
\]
Likewise, for every \(N\ge 1\), if \(\hat\mu_{m,N}^e:=Q_m\hat\pi_{m,N}\), then
\[
\KL(\hat\mu_{m,N}\|\hat\mu_{m,N}^e)
=
\mathbb{E}_{\hat\pi_{m,N}}\bigl[\log d_m(Y)\bigr].
\]
\end{corollary}

\begin{proof}
Corollary~\ref{cor:sturmian-exact-relabeling} gives
\[
H(\mu_m)=H(\pi_m^{\mathrm{fold}}),
\qquad
H(\hat\mu_{m,N})=H(\hat\pi_{m,N}).
\]
Applying Proposition~\ref{prop:fold-entropy-ledger} to \(\mu_m\) and
\(\hat\mu_{m,N}\) yields the two identities.
\end{proof}

A complementary microstate lift formalism is collected in
Appendix~\ref{app:fiber-lift}. It records the fiber-uniform lift family, the
exact entropy ledger used in Corollary~\ref{cor:sturmian-pure-degeneracy-ledger},
and an optional Pythagorean identity for relative entropy against lifted macro
models.
