\section{Concluding Remarks}
\label{sec:concluding}

The canonical Zeckendorf fold provides a deterministic coarse-graining of
binary rotation words into the golden-mean shift language. The main
contribution of this paper is to show that the resulting divergence from the
Parry measure admits a complete finite-\(m\) analysis in which every quantity
is either closed-form or bounded by explicit constants.

The canonical Sturmian slice is the point at which the picture becomes
especially rigid. There the fold is not only injective on realized blocks; it
is conjugate to the classical higher-block presentation of the Sturmian source.
At the same time, the endpoint structure of the golden-mean Parry cylinders is
simple enough to make the Parry comparison exact.
Theorem~\ref{thm:optimal-injective-relabeling} shows that even the best
injective placement of the realized Sturmian law in \(X_m\) still has
divergence \(m\log\varphi-\log m+O_\alpha(1)\) for bounded-type slopes,
uniformly in the R\'enyi order. The folded law lies within a bounded additive
distance of this optimum.

This finite-\(m\) analysis calibrates a universal phenomenon: for any
primitive shift of finite type with topological entropy \(h\), any
deterministic coarse-graining injective on a subexponential-support source
incurs divergence at rate \(mh-\log m+O(1)\)
(Corollary~\ref{cor:low-complexity-source-barrier}). The golden-mean benchmark
is the unique setting in the Fibonacci family where the universal envelope
becomes exact, the optimizer is explicit, and the Fibonacci weighting is forced
by a contiguous numeration axiom
(Theorem~\ref{thm:fibonacci-weight-uniqueness}).

The limitations are also clear. The exact finite-\(m\) formulas rely on the
endpoint collapse geometry of the golden-mean shift. The injective-placement
problem is asymptotic in general, but the support-constrained Parry
optimization is finite-\(m\) exact for every primitive one-step SFT
(Theorem~\ref{thm:primitive-sft-endpoint-envelope}). What is special about the
golden-mean benchmark is therefore not exact solvability itself, but the
complete collapse of the endpoint spectrum to an explicitly enumerable
three-level structure. Whether analogous elementary closed forms exist for
other shifts of finite type with richer endpoint spectra, or for numeration
systems attached to other Pisot units, remains open. By topological conjugacy
(Theorem~\ref{thm:audit-conjugacy-invariance}), the entire theory transfers
verbatim to \(C^2\) circle diffeomorphisms with irrational rotation number.
